---
title: "3.1 - RADMicrobes_BootCamp"
author: "William Shropshire"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/GitHub/radmicrobes/session3/RScripts/")
```

## Traditional Epidemiology 

The first exercise utilizes a dataset from HistData  
```{r download packages, echo = FALSE}
# Download the package manager tool 'pacman' if not in user environment
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  HistData,
  tidyverse,
  lubridate
)
```

## What if John Snow had 'R'? 
```{r John Snow 1854 London cholera death outbreak data, echo=FALSE}
# Load in dataset from HistData package
data(Snow.dates)

# Observe the first six observations in dataset
head(Snow.dates)

# Use an 'ifelse' statement to label dates red prior to 09/08/1854 and darkgreen for dates from 09/08/1854 to 09/30/1854
Snow.dates$clr <- ifelse(Snow.dates$date < mdy("09/08/1854"), "red", "darkgreen")

# Remove 'NA' values from dataset to avoid warnings
Snow.dates <- na.omit(Snow.dates)

# Plot deaths over time
ggplot(Snow.dates, aes(x = date, y = deaths, color = clr)) +
  geom_line(linewidth = 2) +
  geom_point(size = 3, shape = 16) +
  geom_text(aes(x = mdy("09/08/1854"), y = 40, label = "Pump handle\nremoved Sept. 8"), vjust = -1, hjust = -0.25) + # Adding text
  geom_vline(xintercept = mdy("09/08/1854"), linetype = "dotted", color = "red", linewidth = 1) +  # Adding a vertical dotted line
  scale_color_manual(values = c("red", "darkgreen"), breaks = c("red", "darkgreen"), labels = c("Intact", "Removed")) +
  labs(title = "Cholera Deaths Over Time",
       x = "Date",
       y = "Number of Deaths",
       color = "Pump Status") +
  theme_minimal()

# Save the most recent plot as a pdf in working directory 
ggsave("cholera_deaths_time.pdf")
```

```{r create a fictional 1854 Cholera outbreak dataset, echo=FALSE}
# Create monthly time series
dates <- seq(as.Date("1853-08-01"), as.Date("1855-12-01"), by = "month")

# Expand each date into two rows (O1 and non-O1)
outbreak_data <- expand.grid(date = dates, serotype = c("O1", "non-O1")) %>%
  arrange(date)

# Assign outbreak phases
outbreak_data <- outbreak_data %>%
  mutate(period = case_when(
    date >= as.Date("1854-06-01") & date <= as.Date("1854-08-01") ~ "peak1",
    date >= as.Date("1855-07-01") & date <= as.Date("1855-10-01") ~ "peak2",
    date >= as.Date("1853-08-01") & date <= as.Date("1853-09-01") ~ "early_bump",
    TRUE ~ "baseline"
  ))

# Simulate count and deaths
set.seed(42)
outbreak_data <- outbreak_data %>%
  rowwise() %>%
  mutate(
    count = case_when(
      period %in% c("peak1", "peak2") & serotype == "O1" ~ sample(100:160, 1),
      period %in% c("peak1", "peak2") & serotype == "non-O1" ~ sample(20:40, 1),
      period %in% c("early_bump") & serotype == "O1" ~ sample(40:60, 1),
      period %in% c("early_bump") & serotype == "non-O1" ~ sample(10:20, 1),
      TRUE ~ sample(5:15, 1)
    ),
    deaths = case_when(
      serotype == "O1" & period %in% c("peak1", "peak2") ~ round(count * runif(1, 0.40, 0.60)),
      serotype == "O1" ~ round(count * runif(1, 0.15, 0.30)),
      serotype == "non-O1" ~ round(count * runif(1, 0.01, 0.08))
)
  ) %>%
  ungroup()
```

# What if John Snow had (1) knowledge of germ theory; (2) capacity to culture *Vibrio cholera*; (3) either phenotypic or genotypic data at his disposal?
```{r plot of cholera cases from 1853-08 to 1855-12}
# Collapse serotypes to simulate lack of typing
agg_data <- outbreak_data %>%
  group_by(date) %>%
  summarise(
    count = sum(count),
    deaths = sum(deaths)
  )

ggplot(agg_data, aes(x = date)) +
  geom_bar(aes(y = count), stat = "identity", fill = "gray60") +
  geom_line(aes(y = deaths, color = "Deaths"), linewidth = 1.5, linetype = "dotted") +
  labs(title = "Cholera Cases and Deaths Without Serotype Data",
       x = "Date", y = "Cases") +
  scale_color_manual(values = "red") +
  theme_minimal()
```

# Data stratified by Serotype
```{r analyze fictional dataset, message=FALSE}
# Deaths by date for trendline
deaths_by_date <- outbreak_data %>%
  group_by(date) %>%
  summarise(deaths = sum(deaths, na.rm = TRUE))

ggplot(outbreak_data, aes(x = date)) +
  geom_bar(aes(y = count, fill = serotype), stat = "identity") +
  geom_line(data = deaths_by_date, aes(x = date, y = deaths, color = "Deaths"),
            linewidth = 1.5, linetype = "dotted") +
  labs(title = "Cholera Cases and Deaths With Serotype Data",
       x = "Date", y = "Cases") +
  scale_fill_manual(values = c("O1" = "blue", "non-O1" = "green")) +
  scale_color_manual(values = c("Deaths" = "red"), name = "") +
  theme_minimal()

#ggsave("fictional_hist_w_serotype.pdf")

```

```{r Serotype Associated with Cholera attributable mortality}
# Create summary of deaths and survivors per serotype
contingency_table <- outbreak_data %>%
  group_by(serotype) %>%
  summarise(
    Died = sum(deaths, na.rm = TRUE),
    Survived = sum(count - deaths, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  column_to_rownames("serotype") %>%
  as.matrix()

# Run chi-square test
chisq.test(contingency_table)
```
### There is a strong association of Vibrio cholorae serotype 01 attributable mortallity compared to non-01 serotypes (P-value = 2.2e-16). When looking at the mortality over our timeframe stratified by serotype, 01 serotypes had 30% mortality compared to 4% for non-01 Vibrio cholorae.

```{r clean up, echo=FALSE, message=FALSE}
# Clear data
rm(list = ls())  

# Clear packages
pacman::p_unload(all)  

# Clear plots
graphics.off()  

# Clear console
cat("\014") 
```